Promise and peril: Agnostic identification methods\\ for detecting differential item functioning

### 2020-03-23 paper/paper.tex
finished paper print 6 and shipped back to ben

\begin{document}
\maketitle
\begin{abstract}
It is well known that likelihood ratio tests (LRT) are effective at detecting differential item functioning (DIF) in item response models. However, to use a LRT, we require an identifying assumption to disentangle differences in group ability from potential DIF. We use the term ``agnostic identification'' (AI) to describe the process of finding such identifying assumptions without a priori knowledge of relative group ability or items that are DIF-free. We first summarize existing AI methods and propose a variety of new methods. We then conduct a simulation study---which we argue is more realistic than most DIF simulation studies in the literature---and find that one of the proposed new AI methods, all-others-as-anchors-one-at-a-time (AOAA-OAT), significantly outperforms current methods. We also suggest a new approach, the equal means, multiple imputation logit graph (EM-MILG), which clearly presents all information about possible DIF, including sampling variability in item parameters, to the analyst. \clearpage
\end{abstract}

@@

\hypertarget{ben-d-notes}{%
\section{Ben D Notes}\label{ben-d-notes}}

\begin{itemize}
\item
  Switch 1 and 2 to theta and eta
\item
  Ben worries that in anchor items in particular having the demonstration data and the the thought experiment is too many threads
\item
  Be clear upfront that there are two ways to identify: anchor items and anchor points (and be sure to define both)
\item
  Explain AOAA-OAT inuition better. Perhaps using an EM-MILG or something? (in reference to the \enquote{least homogeneous in the sense of} part)
\item
  Compare performance across all methods at once, not one at a time
\item
  Show a paneled MILG for every method all at once. How cool.
\item
  About the simulation difmap figure: \enquote{It seems to me like more could be set about this. you might want to note, for example, that it need not be the case in general that we have 6 items \enquote{clustering} around the same location of the scale. you could have items 1-6 spread out in same way as items 7-12. much harder in that case. in fact, it seems like there is a geometry problem here we could attempt to solve. there are some item-easiness configureations (in sense of MILG) that we really are probably in bad shape with. if the item easiness params are uniformly distributed between -1 and 1 for example, we're stuck! maybe we could try to think about how to characterize such cases and write about what to do there?}
\item
  Fix that equation to take into account sample sizes
\end{itemize}

@@

Following Camilli (\protect\hyperlink{ref-camilli1992conceptual}{1992}), we conceptualize differential item functioning (DIF) as a varying relationship between ability and probability of correct response for students from different groups that manifests when one imposes an item response model with too few ability dimensions. From this perspective, the term \enquote{differential item functioning} is, perhaps, a misnomer as DIF is better thought of as a property of the student, as opposed to the item. For example, Ackerman (\protect\hyperlink{ref-ackerman1992didactic}{1992}) describes a scenario in which a test intends to measure a student's math ability, but performance also depends on their verbal ability. In this case, math ability is the \enquote{target ability,} and verbal ability is the \enquote{nuisance ability.} Fitting a unidimensional item response model to this test results in students with low verbal ability receiving a score systematically lower than their true math ability; therein lies \enquote{DIF}.

Despite this, the usual setup of DIF simulation studies frames DIF as a property of the item. For example, Kopf, Zeileis, and Strobl (\protect\hyperlink{ref-kopf2015framework}{2015}) simulate students as belonging to either a reference or focal group. They fix the item easinesses for the reference group, \({b_j}^{\text{ref}}\), to values that they obtained from a previous study. They set item easinesses for the focal group to \({b_j}^{\text{foc}} = {b_j}^{\text{ref}}\) for items without DIF and \({b_j}^{\text{foc}} = {b_j}^{\text{ref}} - 0.6\) for items with DIF, where 0.6 is the magnitude of DIF in logits. They then simulate student ability \({\theta_i}^{\text{ref}} \sim N(0,1)\) for students in the reference group and \({\theta_i}^{\text{foc}} \sim N(-1,1)\) for students in the focal group. Finally, they generate item responses according to the Rasch model, which specifies that the probability of student \(i\) responding correctly to item \(j\) is
\begin{align}
P(y_{ij} = 1 | \theta_i, b_j) = \sigma(\theta_i + b_j) \label{zopf}
\end{align}
where \(\sigma(x) = \frac{e^x}{e^x + 1}\) is the standard logistic function.

For every DIF simulation study framed in terms of item parameters that vary across groups, there is a mathematically equivalent setup in which students' abilities are multidimensional. For example, to translate the Kopf, Zeileis, and Strobl (\protect\hyperlink{ref-kopf2015framework}{2015}) simulation from the DIF-as-item-property to the DIF-as-student-property view, item easiness is set to what was previously \({b_j}^{\text{ref}}\) for all students. Student ability is expanded to two dimensions, the target ability dimension and nuisance ability dimension. The target ability is the same as unidimensional ability above where \({\theta_i}^{\text{ref}} \sim N(0,1)\) for students in the reference group and \({\theta_i}^{\text{foc}} \sim N(-1,1)\) for students in the focal group. The nuisance ability takes one of two values: \({\eta_i}^{\text{ref}} = 0\) for students in the reference group and \({\eta_i}^{\text{foc}} = -1\) for students in the focal group. We then reply on a 2PL compensatory 2PL model; slopes on the target ability are \(a_{1j} = 1\) for all items (consistent with the Rasch model) and the slope on nuisance ability is \(a_{2j} = 0.6\) for all items with DIF, and \(a_{2j} = 0\) otherwise. Again, 0.6 is the magnitude of DIF in logits. According to the two-dimensional compensatory 2PL model (Thissen and Steinberg \protect\hyperlink{ref-thissen1986taxonomy}{1986}), the probability that student \(i\) responds correctly to item \(j\) is
\begin{align}
\text{Pr}(y_{ij} = 1 | \theta_i, \eta_i, a_{1j}, a_{2j}, b_j) = \sigma(a_{1j}\theta_i + a_{2j}\eta_i + b_j).
\end{align}
For a choice of student (from the focal or reference group) and item (with DIF or without), this model produces identical probabilities as Eqn \ref{zopf}. This translation between views makes explicit that nearly all DIF simulation studies have, perhaps suboptimally, examined the unrealistic scenario in which there is no variation in the nuisance ability for students in the same group.

@@

Psychometricians have long been in search of the perfect method for detecting differential functioning of \emph{individual} items. The IRT-based likelihood ratio test (LRT) is effective at detecting DIF (Meade and Wright \protect\hyperlink{ref-meade2012solving}{2012}). Given both the efficacy or the LRT approach and our preference for methods that existin within the IRT-framework, we do not discuss methods like the Mantel-Haenszel procedure (Holland and Thayer \protect\hyperlink{ref-holland1986differential}{1986}) (which, for example, has been shown to perform no better---and in some cases worse---than IRT-based methods (Swaminathan and Rogers \protect\hyperlink{ref-swaminathan1990detecting}{1990})).

However, use of LRT requires an identifying assumption. The assumption is needed so as to link groups. In the typical setting, no a prior assumptions---about either relative group ability or which items might have DIF---can be made. Thus, we're interested in \emph{agnostic identification} methods (hereafter referred to as \enquote{AI methods}). For simplicity, we focus on the Rasch model to isolate the fundamental issues in \enquote{the AI problem}. Further, so as to offer a coherent framework, we sometimes edit names of existing methods.\footnote{We recognize that others have done the same (e.g., Kopf, Zeileis, and Strobl \protect\hyperlink{ref-kopf2015framework}{2015}), and that we risk contributing to a proliferation of names.}

In this section, we summarize existing AI methods and propose a variety of extensions. We use a (simulated) demonstration dataset to illustrate key features of the different methods.\footnote{This data is used solely for illustrative purposes; we describe an extensive simulation study later in the paper.} The data consists of 10,000 reference group students and 10,000 focal group students taking an eight-item test. Our simulation is similar to our multidimensional alternative to the Kopf, Zeileis, and Strobl (\protect\hyperlink{ref-kopf2015framework}{2015}) approach discussed above. Target ability is simulated as \({\theta_i}^{\text{ref}} \sim N(\mu^\text{ref} = 0,1)\) for students in the reference group and \({\theta_i}^{\text{foc}} \sim N(\mu^\text{foc} = -1,1)\) for students in the focal group. Nuisance ability is set to \({\eta_i}^{\text{ref}} = 0\) and \({\eta_i}^{\text{foc}} = -1\). The slope on target ability is set to \(a_{1j} = 1\) for all items. The slope on nuisance ability is set to \(a_{2j} = 0.5\) for the last three items (the items with DIF) and \(a_{2j} = 0\) otherwise. We can also, of course, describe these conditions from the DIF-as-item-property view where there is no nuisance ability. Instead, \({b_j}^{\text{foc}} = {b_j}^{\text{ref}}\) for the first five items and \({b_j}^{\text{foc}} = {b_j}^{\text{ref}} - 0.5\) for the last three items. To be sure, this data is used only for demonstrative purposes. We compare the performance of the AI methods in a more realistic scenario in the \protect\hyperlink{simstudy}{the simulation study}.

The fundamental challenge is to use an AI method to disentangle estimation of \(\hat\mu^\text{foc} - \mu^\text{ref}\), the difference in mean abilities, from the estimation of \(\hat d = \hat{b_j}^{\text{foc}} - \hat{b_j}^{\text{ref}}\), the difference in item easiness. The most common AI approaches identify a group of anchor items that are assumed to be DIF-free. These anchor items identify the model, thereby allowing for the estimation of \(\hat\mu^\text{foc}\), and the remaining items can be tested for DIF using an LRT.

Throughout this paper, we estimate models using marginal maximum likelihood estimation (Bock and Aitkin \protect\hyperlink{ref-bock1981marginal}{1981}). In addition, we follow the common and inconsequential practice of identifying the scale by setting \(\mu^\text{ref} = 0\).

@@

While we will document the performance of these methods momentarily, we first suggest an approach focused on putting analysts in a better position for understanding the scale of the problem in their data. We propose the \enquote{equal means, multiple imputation logit graph} (EM-MILG), which presents information about potential DIF to the analyst. EM-MILG begins by fitting a unidimensional Rasch model to the data that is identified by arbitrarily setting \(\mu^\text{foc} = 0\). Given that we also assume \(\mu^\text{ref} = 0\), this is equivalent to the assumption that the groups have equal mean ability. As a result, all differences in performance---either from group ability differences or DIF---manifest in the item easiness difference parameter \(\tilde{d_j}\). The tilde (e.g., \(\tilde{d_j}\)) is used to denote parameters estimated with \(\mu^\text{foc}\) set to \(0\)---as compared to the hat (e.g., \(\hat{d_j}\)), which is used for parameter estimates from the properly identified model.

@@

Figure \ref{fig:emmilg} shows the EM-MILG for our demonstration data. We emphasize that the EM-MILG contains all possible information about the difference in group performance. The analyst might assume that---because there are five items where the reference group outperforms the focal group by approximately 1 logit and only three items where the difference is 1.5 logits---items 1-5 are unbiased. Indeed, all AI methods rest on the logic that its the minority of items that contain DIF. These unbiased items are known as anchor items; setting anchor items is the most common identification assumption. After anchor items have been selected, the model is refit where the assumption that \(\mu^\text{foc} = 0\) is dropped and replaced with the assumption that the anchor items are DIF-free (i.e., \(d_j = 0\)). Of course, the demonstration data is designed such that there are obvious groups; this will not typically be the case.

@@

The same process of using multiple imputations to estimate the distribution of \(\tilde{d_j}\) can be used with the final model. Because the equal means assumption is not made, we refer to the resulting visualization as a multiple imputations logit graph (MILG). As expected, in our demonstration data, selecting the first five items as anchors correctly results in \(\hat d_j \approx 0.5\) for the items with DIF as is shown in the MILG in Figure \ref{fig:milg}.

@@

In general, one of our key concerns with traditional AI methods is that they can lull the analyst into a false sense of security. The analyst simply chooses a method, implements it, and then proceeds as if the method certainly resulted in a model that reflects some objective reality. Our aim with EM-MILG is to combat this concern by presenting all information clearly to the analyst. In the previous example, the analyst may be wary of their results, having seen how arbitrary it was to conclude that the first five and not the last three items are unbiased. Even when other AI methods are used, the analyst can use the EM-MILG as the first step in order to give them a sense of their item response data. And, of course, the MILG can be used to visualize DIF regardless of how the model is identified. We now move on to more traditional AI methods, which don't require judgment from the analyst.

@@

Meade and Wright (\protect\hyperlink{ref-meade2012solving}{2012}) compared the most commonly used AI methods and unequivocally recommended the all-others-as-anchors (AOAA) method. AOAA tests each item for DIF one at a time using all of the other items as anchors. For example, when testing the first item for DIF, all of the other items are used as anchors. This is done using a LRT that compares the baseline model, where all item parameters are fixed across groups, to the flexible model, where the parameters of the tested item are freed across groups (Thissen, Steinberg, and Wainer \protect\hyperlink{ref-thissen1993detection}{1993}). Then, when testing the second item for DIF, once again all of the other items (including the first item) are used as anchors, and so on. The items for which the flexible model outperforms the baseline model (typically based on a \(\chi^2\) test) are identified as having DIF, and the rest of the items become anchor items. AOAA is implemented in the mirt R package, and is called by passing scheme = \enquote{drop} to the DIF function (drop refers to dropping a single constraint when moving from the constrained to the flexible model).

Implicit in the use of AOAA is the assumption that all items not being tested do not exhibit DIF, which is, of course, counter to the underlying rationale for the undertaking. On a practical level, it is thought that AOAA will perform well if a small minority of items have DIF or the DIF is balanced such that some items are biased against the focal group, while others are biased against the reference group.

Researchers have noticed the circular logic of AOAA, but have mostly described it indirectly by pointing out inflated Type I errors in simulation studies (Stark, Chernyshenko, and Drasgow \protect\hyperlink{ref-stark2006detecting}{2006}). A simple thought experiment illustrates how AOAA fails: Imagine a three item test with a sufficiently large number of students where the first item has DIF, and the other two do not. Using AOAA, all items test positive for DIF. The last two items incorrectly test positive because including the first item in the anchor set causes the group ability difference to be misestimated. This phenomenon of items with real DIF inducing the appearance of DIF in other items was only indirectly discussed in the literature until Andrich and Hagquist (\protect\hyperlink{ref-andrich2012real}{2012}) coined the term \enquote{artificial DIF}, which is a significant problem in applications of AI methods as discussed below.

@@

One way to attempt to counter artificial DIF is with purification through iterative anchor selection. For example, Drasgow (\protect\hyperlink{ref-drasgow1987study}{1987}) started with AOAA, removed items displaying DIF from the anchor set, then repeated the process iteratively---with items that have been removed from the anchor set allowed to have free parameters across groups in both the baseline and flexible model---until no more items tested positively. Kopf, Zeileis, and Strobl (\protect\hyperlink{ref-kopf2015framework}{2015}) named this technique Iterative-backward-AOAA with \enquote{backward} (as in reverse) referring to beginning with the assumption that all items are DIF-free. We find it clearer to refer to this method as all-others-as-anchors-all-significiant (AOAA-AS). Appending \enquote{all-significant} indicates that anchor selection is made iteratively with all items that test positive for DIF being removed from the anchor set. AOAA-AS is implemented in the mirt R package, and is called by passing scheme = \enquote{drop\_sequential} to the DIF function.

We argue that AOAA-AS, while a potential improvement, doesn't solve the fundamental problem of AOAA: What does one do when all items test positive for DIF? With a sufficient sample size and at least one item with DIF, this will necessarily be the case. In our thought experiment, we get the same result with AOAA-AS as we did with AOAA: All items test positive for DIF, and there are no anchor items. Kopf, Zeileis, and Strobl (\protect\hyperlink{ref-kopf2015framework}{2015}) encountered precisely this problem in their simulation study and chose to select a single anchor item randomly. Woods (\protect\hyperlink{ref-woods2009empirical}{2009}) suggested a more straightforward, one-step method: Use AOAA and select the four items that exhibit the least amount of DIF as anchor items. It's unclear if one should proceed if those four items also test positive for DIF.

@@

We propose an extension of these methods, all-others-as-anchors-one-at-a-time (AOAA-OAT), which, to our knowledge (and surprise), has not previously been explicitly proposed. AOAA-OAT is inspired by Hagquist and Andrich (\protect\hyperlink{ref-hagquist2017recent}{2017}), who assert that \enquote{items showing DIF initially should not be resolved simultaneously but sequentially} {[}p.~6{]}. Like AOAA-AS, AOAA-OAT starts with AOAA, but only the single item exhibiting the most DIF, based on the \(\chi^2\) test statistic, is removed from the anchor set. The process continues iteratively until no new items display DIF. AOAA-OAT and AOAA-AS are similar in that they are both iterative; the difference is that AOAA-OAT takes the more conservative approach of removing only \emph{one} item in each iteration as opposed to \emph{all} items that test positive for DIF. As a result, we believe that AOAA-OAT is less likely to be \enquote{fooled} by artifical DIF. AOAA-OAT is not currently implemented in the R package mirt.

Applying AOAA-OAT to our thought experiment demonstrates its effectiveness. The initial AOAA removes the real DIF item from the anchor set because it exhibits the most DIF (i.e., it is the least homogeneous in the sense that the group difference in performance is most different from the average group difference in performance on the other two items). In the next step, both of the other items test negative for DIF, and we arrive at the correct conclusion. To work, AOAA-OAT has two requirements: First, that at least two items do not have DIF, and second, that the set of items without DIF are more homogeneous than other sets of items.

\hypertarget{summary-and-performance}{%
\paragraph{Summary and performance}\label{summary-and-performance}}

It's useful to remember that AOAA is not an iterative procedure. In constrast, the methods with a hyphen, AOAA-OAT and AAOA-AS, are iterative procedures with AOAA-OAT being the new, more conservative method. All of the methods described in this paper are summarized in Table \ref{table:allmethods}. For our demonstration data, AOAA and AOAA-AS both failed to select any anchor items, leaving the model unidentified. On the other hand, AOAA-OAT worked perfectly, selecting exactly the first five items as anchor items.

@@

Bechger and Maris (\protect\hyperlink{ref-bechger2015statistical}{2015}) proposed selecting anchor items by identifying clusters of items that function similarly and then choosing one of those clusters to be the \enquote{anchor cluster}. They pointed out that one way around the unidentifiability issue is to consider only relative item parameters. For each group, the relative easinesses for each pair of items can be stored in the (square) matrix \(\mathbf{R}^{\text{ref}}\) with entries \({R_{xy}}^{\text{ref}} = {b_x}^{\text{ref}} - {b_y}^{\text{ref}}\), where \(x\) and \(y\) both index the items. A similar matrix, \(\mathbf{R}^{\text{foc}}\), can be formed. The ultimate matrix of interest is \(\Delta \mathbf{R} \equiv \mathbf{R}^{\text{ref}}-\mathbf{R}^{\text{foc}}\) which is the \enquote{differences between groups in the pairwise differences in (easiness) between items} {[}p.~323{]}.

>>>>

\begin{document}
\maketitle
\begin{abstract}
It is well known that likelihood ratio tests (LRT) are effective at detecting differential item functioning (DIF) in item response models. However, to use a LRT (or any other method), we require an identifying assumption to disentangle differences in group ability from potential DIF. We use the term ``agnostic identification'' (AI) to describe the process of finding such identifying assumptions without a priori knowledge of relative group ability or items that are DIF-free. We first summarize existing AI methods and propose a variety of new methods. We then conduct a simulation study---which we argue is more realistic than most DIF simulation studies in the literature---and find that one of the proposed new AI methods, all-others-as-anchors-one-at-a-time (AOAA-OAT), significantly outperforms current methods. We also suggest a new approach, the equal means, multiple imputation logit graph (EM-MILG), which clearly presents all information about possible DIF, including sampling variability in item parameters, to the analyst. \clearpage
\end{abstract}

@@

@@

Following Camilli (\protect\hyperlink{ref-camilli1992conceptual}{1992}), we conceptualize differential item functioning (DIF) as a varying relationship between ability and probability of correct response for students from different groups that manifests when one imposes an item response model with too few ability dimensions. From this perspective, the term \enquote{differential item functioning} is, perhaps, a misnomer as DIF is better thought of as a property of the student, as opposed to the item. For example, Ackerman (\protect\hyperlink{ref-ackerman1992didactic}{1992}) describes a scenario in which a test intends to measure a student's math ability, but performance also depends on their verbal ability. In this case, math ability is the \enquote{target ability}, and verbal ability is the \enquote{nuisance ability}. Fitting a unidimensional item response model to this test results in students with low verbal ability receiving a score systematically lower than their true math ability; therein lies \enquote{DIF}.

Despite this, the usual setup of DIF simulation studies frames DIF as a property of the item. For example, Kopf, Zeileis, and Strobl (\protect\hyperlink{ref-kopf2015framework}{2015}) simulate students as belonging to either a reference or focal group. They fix the item easinesses for the reference group, \({b_j}^{\text{ref}}\), to values that they obtained from a previous study. They set item easinesses for the focal group to \({b_j}^{\text{foc}} = {b_j}^{\text{ref}}\) for items without DIF and \({b_j}^{\text{foc}} = {b_j}^{\text{ref}} - 0.6\) for items with DIF, where 0.6 is the magnitude of DIF in logits. They then simulate student ability \({\theta_i}^{\text{ref}} \sim N(0,1)\) for students in the reference group and \({\theta_i}^{\text{foc}} \sim N(-1,1)\) for students in the focal group. Finally, they generate item responses according to the Rasch model, which specifies that the probability of student \(i\) responding correctly to item \(j\) is
\begin{align}
\text{Pr}(y_{ij} = 1 | \theta_i, b_j) = \sigma(\theta_i + b_j) \label{zopf}
\end{align}
where \(\sigma(x) = \frac{e^x}{e^x + 1}\) is the standard logistic function.

For every DIF simulation study framed in terms of item parameters that vary across groups, there is a mathematically equivalent setup in which students' abilities are multidimensional. According to the two-dimensional compensatory 2PL model (Thissen and Steinberg \protect\hyperlink{ref-thissen1986taxonomy}{1986}), the probability that student \(i\) responds correctly to item \(j\) is
\begin{align}
\text{Pr}(y_{ij} = 1 | \theta_i, \eta_i, a_{\theta j}, a_{\eta j}, b_j) = \sigma(a_{\theta j}\theta_i + a_{\eta j}\eta_i + b_j).
\end{align}
where \(\theta\) is target ability, \(\eta\) is nuisance ability, \(a_{\theta j}\) is an item's slope on target ability, and \(a_{\eta j}\) is an item's slope on nuisance ability.

For example, to translate the Kopf, Zeileis, and Strobl (\protect\hyperlink{ref-kopf2015framework}{2015}) simulation from the DIF-as-item-property to the DIF-as-student-property view, \(b_j\) is set to \({b_j}^{\text{ref}}\), \(\theta\) stays the same, \(a_{\theta j} = 1\) for all items (consistent with the Rasch model), and \(\eta\) takes one of two values: \({\eta_i}^{\text{ref}} = 0\) for students in the reference group and \({\eta_i}^{\text{foc}} = -1\) for students in the focal group. For items with DIF, \(a_{\eta j} = 0.6\), and for items without DIF, \(a_{\eta j} = 0\) Again, 0.6 is the magnitude of DIF in logits.\footnote{The reader may notice that this is just one possible specification. Any specification that meets the following criteria has the same effect: First, \(a_{\eta j}\eta_i\) must be 0.6 lower for the focal group on items with DIF. Second, \(a_{\eta j}\eta_i\) must be equal across groups for items without DIF.} For a choice of student (from the focal or reference group) and item (with DIF or without), this model produces identical probabilities as Eqn \ref{zopf}. This translation between views makes explicit that nearly all DIF simulation studies have, perhaps suboptimally, examined the unrealistic scenario in which there is no variation in the nuisance ability for students in the same group.

@@

Psychometricians have long been in search of the perfect method for detecting differential functioning of \emph{individual} items. The IRT-based likelihood ratio test (LRT) is effective at detecting DIF (Meade and Wright \protect\hyperlink{ref-meade2012solving}{2012}). Given both the efficacy or the LRT approach and our preference for methods that existin within the IRT-framework, we do not discuss methods like the Mantel-Haenszel procedure (Holland and Thayer \protect\hyperlink{ref-holland1986differential}{1986}), which has been shown to perform no better---and in some cases worse---than IRT-based methods (Swaminathan and Rogers \protect\hyperlink{ref-swaminathan1990detecting}{1990})). Within the IRT-framework, we---as is common---estimate models using marginal maximum likelihood estimation (Bock and Aitkin \protect\hyperlink{ref-bock1981marginal}{1981}) and identify the scale by setting \(\mu^\text{ref} = 0\).

Use of LRT requires an identifying assumption. The assumption is needed so as to link groups. In the typical setting, no a prior assumptions---about either relative group ability or which items might have DIF---can be made. Thus, we're interested in \emph{agnostic identification} methods (hereafter referred to as \enquote{AI methods}). The fundamental challenge is to disentangle estimation of \(\hat\mu^\text{foc} - \mu^\text{ref} = \hat\mu^\text{foc} - 0 = \hat\mu^\text{foc}\), the difference in mean abilities, from the estimation of \(\hat d = \hat{b_j}^{\text{foc}} - \hat{b_j}^{\text{ref}}\), the difference in item easiness. There are two types of AI methods: The most common, the use of anchor items, selects a group of items that are assumed to be DIF-free. These anchor items identify the model, thereby allowing for the estimation of \(\hat\mu^\text{foc}\), and the remaining items can be tested for DIF using an LRT. The other type, the use of an anchor point (Strobl et al. \protect\hyperlink{ref-strobl2018anchor}{2018}), directly sets the difference in mean abilities to some value, \(\mu^{\star\text{foc}}\). This anchor point identifies the model, and all items can be tested for DIF using an LRT.

In this section, we summarize existing AI methods and propose a variety of extensions. For simplicity, we focus on the Rasch model to isolate the fundamental issues in \enquote{the AI problem}. Further, so as to offer a coherent framework, we sometimes edit names of existing methods.\footnote{We recognize that others have done the same (e.g., Kopf, Zeileis, and Strobl \protect\hyperlink{ref-kopf2015framework}{2015}), and that we risk contributing to a proliferation of names.} We use a (simulated) demonstration dataset to illustrate key features of the different methods.\footnote{This data is used solely for illustrative purposes; we describe an extensive simulation study later in the paper.} The data consists of 10,000 reference group students and 10,000 focal group students taking an eight-item test. Our simulation is similar to our multidimensional alternative to the Kopf, Zeileis, and Strobl (\protect\hyperlink{ref-kopf2015framework}{2015}) approach discussed above. Target ability is simulated as \({\theta_i}^{\text{ref}} \sim N(\mu^\text{ref} = 0,1)\) for students in the reference group and \({\theta_i}^{\text{foc}} \sim N(\mu^\text{foc} = -1,1)\) for students in the focal group. Nuisance ability is set to \({\eta_i}^{\text{ref}} = 0\) and \({\eta_i}^{\text{foc}} = -1\). The slope on target ability is set to \(a_{1j} = 1\) for all items. The slope on nuisance ability is set to \(a_{2j} = 0.5\) for the last three items (the items with DIF) and \(a_{2j} = 0\) otherwise. We can also, of course, describe these conditions from the DIF-as-item-property view where there is no nuisance ability. Instead, \({b_j}^{\text{foc}} = {b_j}^{\text{ref}}\) for the first five items and \({b_j}^{\text{foc}} = {b_j}^{\text{ref}} - 0.5\) for the last three items. To be sure, this data is used only for demonstrative purposes. We compare the performance of the AI methods in a more realistic scenario in the \protect\hyperlink{simstudy}{the simulation study}.

The AI methods described in this paper are summarized in Table \ref{table:allmethods}. In addition the results for each method on the demonstration data are shown in a MILG in Figure \ref{fig:allmethodsfig}.

@@

While we will document the performance of these methods momentarily, we first suggest an approach focused on putting analysts in a better position to understand the scale of the problem in their data. We propose the \enquote{equal means, multiple imputation logit graph} (EM-MILG), which presents information about potential DIF to the analyst. EM-MILG begins by fitting a unidimensional Rasch model to the data that is identified by arbitrarily setting \(\mu^\text{foc} = 0\). Given that we also assume \(\mu^\text{ref} = 0\), this is equivalent to the assumption that the groups have equal mean ability. As a result, all differences in performance---either from group ability differences or DIF---manifest in the item easiness difference parameter \(\tilde{d_j}\). The tilde (e.g., \(\tilde{d_j}\)) is used to denote parameters estimated with \(\mu^\text{foc}\) set to \(0\)---as compared to the hat (e.g., \(\hat{d_j}\)), which is used for parameter estimates from a properly identified model.

@@

Figure \ref{fig:emmilg} shows the EM-MILG for our demonstration data. We emphasize that the EM-MILG contains all possible information about the difference in group performance. The analyst might assume that---because there are five items where the reference group outperforms the focal group by approximately 1 logit and only three items where the difference is 1.5 logits---items 1-5 are unbiased. This might seem to be a strong assumption, but all AI methods assume---in one way or another---that, if there is DIF, its a minority of items that are responsible. These unbiased items are known as anchor items; setting anchor items is the most common identification assumption. After anchor items have been selected, the model is refit where the identifying assumption that \(\mu^\text{foc} = 0\) is dropped and replaced with the identifying assumption that the anchor items are DIF-free (i.e., \(d_j = 0\) for anchor items). Of course, the demonstration data is designed such that there are obvious groups; this will not necessarily typically be the case.

@@

The same process of using multiple imputations to estimate the distribution of \(\tilde{d_j}\) can be used with a properly identified model. Because the equal means assumption is not made, we refer to the resulting visualization as a multiple imputations logit graph (MILG). As expected, in our demonstration data, selecting the first five items as anchors correctly results in \(\hat d_j \approx 0.5\) for the items with DIF as is shown in the MILG in Figure \ref{fig:milg}.

@@

In general, one of our main concerns with traditional AI methods is that they can lull the analyst into a false sense of security: The analyst chooses a method, implements it, and then proceeds as if the method certainly resulted in a model that reflects some objective reality. Our aim with EM-MILG is to combat this concern by presenting all information clearly to the analyst. In the previous example, the analyst may be wary of their results, having seen how arbitrary it was to conclude that the first five and not the last three items are unbiased. Even when other AI methods are used, the analyst can use the EM-MILG as the first step in order to give them a sense of their item response data. And, of course, the MILG can be used to visualize DIF regardless of how the model is identified. We now move on to more traditional AI methods, which don't require judgment from the analyst, but make the majority-of-items-don't-have-DIF assumption more subtly.

@@

Meade and Wright (\protect\hyperlink{ref-meade2012solving}{2012}) compared the most commonly used AI methods and unequivocally recommended the all-others-as-anchors (AOAA) method. AOAA tests each item for DIF one at a time using all of the other items as anchors. For example, when testing the first item for DIF, all of the other items are used as anchors. This is done using a LRT that compares the baseline model, where all item parameters are fixed across groups, to the flexible model, where the parameters of the tested item are freed across groups (Thissen, Steinberg, and Wainer \protect\hyperlink{ref-thissen1993detection}{1993}). Then, when testing the second item for DIF, once again all of the other items (including the first item) are used as anchors, and so on. The items for which the flexible model outperforms the baseline model (typically based on a \(\chi^2\) test) are identified as having DIF, and the rest of the items become anchor items. AOAA is implemented in the mirt R package, and is called by passing scheme = \enquote{drop} to the DIF function (drop refers to dropping a single constraint when moving from the baseline to the flexible model).

Implicit in the use of AOAA is the assumption that all items not being tested do not exhibit DIF, which is, of course, counter to the underlying rationale for the undertaking. On a practical level, it is thought that AOAA will perform well if a small minority of items have DIF or the DIF is balanced such that some items are biased against the focal group, while others are biased against the reference group. Researchers have noticed the circular logic of AOAA, but have mostly described it indirectly by pointing out inflated Type I errors in simulation studies (Stark, Chernyshenko, and Drasgow \protect\hyperlink{ref-stark2006detecting}{2006}). A simple thought experiment illustrates how AOAA fails: Imagine a three item test with a sufficiently large number of students where the first item has DIF, and the other two do not. Using AOAA, all items test positive for DIF. The last two items incorrectly test positive because including the first item in the anchor set causes the group ability difference to be misestimated. This phenomenon of items with real DIF inducing the appearance of DIF in other items was only indirectly discussed in the literature until Andrich and Hagquist (\protect\hyperlink{ref-andrich2012real}{2012}) coined the term \enquote{artificial DIF}, which is a significant problem in applications of AI methods.

@@

One way to attempt to counter artificial DIF is with purification through iterative anchor selection. For example, Drasgow (\protect\hyperlink{ref-drasgow1987study}{1987}) started with AOAA, removed items displaying DIF from the anchor set, then repeated the process iteratively---with items that have been removed from the anchor set allowed to have free parameters across groups in both the baseline and flexible model---until no more items tested positively. Kopf, Zeileis, and Strobl (\protect\hyperlink{ref-kopf2015framework}{2015}) named this technique Iterative-backward-AOAA with \enquote{backward} (as in reverse) referring to beginning with the assumption that all items are DIF-free. We find it clearer to refer to this method as all-others-as-anchors-all-significiant (AOAA-AS). We append \enquote{all-significant} to indicate that all items that test positive for DIF are removed from the anchor set in each iteration. AOAA-AS is implemented in the mirt R package, and is called by passing scheme = \enquote{drop\_sequential} to the DIF function.

We argue that AOAA-AS, while a potential improvement, doesn't solve the fundamental problem of AOAA: What does one do when all items test positive for DIF? With a sufficient sample size and at least one item with DIF, this will necessarily be the case. In our thought experiment, we get the same result with AOAA-AS as we did with AOAA: All items test positive for DIF, and there are no anchor items. Kopf, Zeileis, and Strobl (\protect\hyperlink{ref-kopf2015framework}{2015}) encountered precisely this problem in their simulation study and chose to select a single anchor item randomly. Woods (\protect\hyperlink{ref-woods2009empirical}{2009}) suggested a more straightforward, one-step method: Use AOAA and select the four items that exhibit the least amount of DIF as anchor items.

@@

We propose an extension of these methods, all-others-as-anchors-one-at-a-time (AOAA-OAT), which, to our knowledge (and surprise), has not previously been explicitly proposed. AOAA-OAT is inspired by Hagquist and Andrich (\protect\hyperlink{ref-hagquist2017recent}{2017}), who assert that \enquote{items showing DIF initially should not be resolved simultaneously but sequentially} {[}p.~6{]}. Like AOAA-AS, AOAA-OAT starts with AOAA, but only the single item exhibiting the most DIF---based on the \(\chi^2\) test statistic---is removed from the anchor set. The process continues iteratively until no new items display DIF. AOAA-OAT and AOAA-AS are similar in that they are both iterative; the difference is that AOAA-OAT takes the more conservative approach of removing only \emph{one} item in each iteration as opposed to \emph{all} items that test positive for DIF. As a result, we believe that AOAA-OAT is less likely to be \enquote{fooled} by artifical DIF. AOAA-OAT is not currently implemented in the mirt R package.

Applying AOAA-OAT to our thought experiment demonstrates its effectiveness. The initial AOAA removes the real DIF item from the anchor set because it exhibits the most DIF (i.e., because the group difference in performance is most different from the average group difference in performance on the other two items). In the next step, both of the other items test negative for DIF, and we arrive at the correct conclusion.

@@

Bechger and Maris (\protect\hyperlink{ref-bechger2015statistical}{2015}) proposed selecting anchor items by identifying clusters of items that function similarly and then choosing one of those clusters to be the \enquote{anchor cluster}. They pointed out that one way around the lack of model identifiability is to consider only relative item parameters. For each group, the relative easinesses for each pair of items can be stored in the (square) matrix \(\mathbf{R}^{\text{ref}}\) with entries \({R_{xy}}^{\text{ref}} = {b_x}^{\text{ref}} - {b_y}^{\text{ref}}\), where \(x\) and \(y\) both index the items. A similar matrix, \(\mathbf{R}^{\text{foc}}\), can be formed. The ultimate matrix of interest is \(\Delta \mathbf{R} \equiv \mathbf{R}^{\text{ref}}-\mathbf{R}^{\text{foc}}\) which is the \enquote{differences between groups in the pairwise differences in (easiness) between items} {[}p.~323{]}.
